ranacapa: An R package and Shiny web app to explore environmental DNA data with exploratory statistics and interactive visualizations

### 2018-10-16 manuscript.Rmd
submission commit!

The targeted amplification and sequencing of DNA that living organisms shed into their physical environment, termed "environmental DNA (eDNA) metabarcoding",  is revolutionizing microbiology, ecology, and conservation research [@Taberlet_2012; @Deiner_2017].  Sequencing of eDNA extracted from field-collected soil, water, or sediment samples can yield insight into a range of questions, from profiling the composition of ancient plant and animal communities [@Pedersen_2014], to monitoring populations of rare or endangered species [@Balasingham_2017]. As the cost of eDNA metabarcoding declines and sample collection techniques become more streamlined (e.g. @Thomas_2018), professional research scientists are increasingly using eDNA metabarcoding as a platform to engage a diversity of community partners, including natural resource managers, undergraduate students, and citizen scientists in primary research. However, developing robust and impactful community science programs that engage community partners in all steps of the research process remains a challenge.

@@

This tab also includes a dendrogram that groups sites based on Ward’s cluster analysis (`stats::hclust(distance_object, method = "ward.d2")`), where `distance_object` is made using `phyloseq::distance()`. For both figures, users can toggle between using Jaccard and Bray-Curtis dissimilarity.

- **Beta diversity statistics**: Shows results from two statistical tests of species turnover across site. The first test is a multivariate ANOVA of taxon turnover across sites, implemented with `vegan::adonis())`. The second statistical test, which is implemented with `vegan::betadisper()`, is of heterogeneity of variances among samples. This test compares the degree of sample-to-sample variation within habitats (or within other user-selected groups).

@@

The taxonomy heatmap of species detected using the 12S and CO1 metabarcodes (Figure 2) was the most valuable visualization to this collaboration, because it allowed the resource managers to filter the large observed species list down to a particular set of key taxa that they regularly monitor. The heatmap showed that this pilot study detected 36 of the 70 key metazoans at the species level, and the remaining 34 at the genus, family, or order level. This indicates that eDNA-based studies can provide critical information for ongoing management efforts and provide new insights into the spatial and temporal distributions of these species. The value of `ranacapa` in this scenario was to quickly sort through long species lists generated in by eDNA sequencing to highlight the strengths and weaknesses in using eDNA to monitor diversity in the Channel Islands. The data from this study are packaged as the demo dataset for the `ranacapa` Shiny app and are available online [@Kandlikar_2018a].

@@

The course instructors used the `ranacapa` Shiny app to introduce students to the structure of eDNA sequencing results. The students were encouraged to explore data and perform the statistical analyses most pertinent to the hypotheses they had formed at the beginning of the course. A key benefit of using `ranacapa` was that despite having no prior bioinformatics experience, students could begin exploring the biodiversity in their samples in a matter of minutes by using the online instance of Shiny app. This allowed the instructors to focus classroom time on biological questions rather than on troubleshooting bioinformatics problems, as had been the case in previous sessions of the course. The course instructors noted that visualizing eDNA data in `ranacapa` helped students understand the relationships between taxon-by-site matrices and the various metadata they had collected in the field. By significantly reducing the time and difficulty in visualizing basic biodiversity patterns, `ranacapa` helped students develop and pursue more sophisticated analyses during the remainder of the course, using tools such as STAMP [@Parks_2014] and PICRUSt [@Langille_2013]. The taxonomy tables and metadata files used in this course are available online [@Kandlikar_2018b].

@@

- A Shiny app, including a dataset generated for demonstrations, is available at https://gauravsk.shinyapps.io/ranacapa  
- `ranacapa` is available for installation at https://github.com/gauravsk/ranacapa  
- Link to source code as at time of publication (F1000Research TO GENERATE)  
- Link to archived source code as at time of publication (F1000Research TO GENERATE)  
- Software license (GPL-3)  
- Data availability   
- Figshare: https://doi.org/10.6084/m9.figshare.7199477 and https://doi.org/10.6084/m9.figshare.7199510.v1  
- License: CC-BY 4.0

@@

## Competing interests
No competing interests were disclosed

@@

#### Figure 3
Taxonomy barplot as shown in the `ranacapa` Shiny app. Taxonomy is shown at the Order level in this figure; in the app, users can choose the taxonomic level to show in the barplot
```{r}
fig3
```

>>>>

The targeted amplification and sequencing of DNA that living organisms shed into their physical environment, termed "environmental DNA (eDNA) metabarcoding,"  is revolutionizing microbiology, ecology, and conservation research [@Taberlet_2012; @Deiner_2017].  Sequencing of eDNA extracted from field-collected soil, water, or sediment samples can yield insight into a range of questions, from profiling the composition of ancient plant and animal communities [@Pedersen_2014], to monitoring populations of rare or endangered species [@Balasingham_2017]. As the cost of eDNA metabarcoding declines and sample collection techniques become more streamlined (e.g. @Thomas_2018), professional research scientists are increasingly using eDNA metabarcoding as a platform to engage a diversity of community partners, including natural resource managers, undergraduate students, and citizen scientists in primary research. However, developing robust and impactful community science programs that engage community partners in all steps of the research process remains a challenge.

@@

This tab also includes a dendrogram that groups sites based on Ward’s cluster analysis (`stats::hclust(distance_object, method = "ward.d2")`), where `distance_object` is made using `phyloseq::distance()`. For both figures, users can toggle between using Jaccard and Bray-Curtis dissimilarity. 
- **Beta diversity statistics**: Shows results from two statistical tests of species turnover across sites. The first test is a multivariate ANOVA of taxon turnover across sites, implemented with `vegan::adonis()`. The second statistical test, which is implemented with `vegan::betadisper()`, is of heterogeneity of variances among samples. This test compares the degree of sample-to-sample variation within habitats (or within other user-selected groups).

@@

The taxonomy heatmap of species detected using the 12S and CO1 metabarcodes (Figure 2) was the most valuable visualization to this collaboration, because it allowed the resource managers to filter the large observed species list down to a particular set of key taxa that they regularly monitor. The heatmap showed that this pilot study detected 36 of the 70 key metazoans at the species level, and the remaining 34 at the genus, family, or order level. This indicates that eDNA-based studies can provide critical information for ongoing management efforts and provide new insights into the spatial and temporal distributions of these species. The value of `ranacapa` in this scenario was to quickly sort through long species lists generated by eDNA sequencing to highlight the strengths and weaknesses in using eDNA to monitor diversity in the Channel Islands. The data from this study are packaged as the demo dataset for the `ranacapa` Shiny app and are available online [@Kandlikar_2018a].

@@

The course instructors used the `ranacapa` Shiny app to introduce students to the structure of eDNA sequencing results. The students were encouraged to explore data and perform the statistical analyses most pertinent to the hypotheses they had formed at the beginning of the course. A key benefit of using `ranacapa` was that despite having no prior bioinformatics experience, students could begin exploring the biodiversity in their samples in a matter of minutes by using the online instance of the Shiny app. This allowed the instructors to focus classroom time on biological questions rather than on troubleshooting bioinformatics problems, as had been the case in previous sessions of the course. The course instructors noted that visualizing eDNA data in `ranacapa` helped students understand the relationships between taxon-by-site matrices and the various metadata they had collected in the field. By significantly reducing the time and difficulty in visualizing basic biodiversity patterns, `ranacapa` helped students develop and pursue more sophisticated analyses during the remainder of the course, using tools such as STAMP [@Parks_2014] and PICRUSt [@Langille_2013]. The taxonomy tables and metadata files used in this course are available online [@Kandlikar_2018b].

@@

- A Shiny app, including a dataset generated for demonstrations, is available at https://gauravsk.shinyapps.io/ranacapa  
- `ranacapa` is available for installation at https://github.com/gauravsk/ranacapa  
- Link to source code as at time of publication (F1000Research TO GENERATE)  
- Link to archived source code as at time of publication (F1000Research TO GENERATE)  
- Software license (GPL-3)  
- Data available on Figshare: https://doi.org/10.6084/m9.figshare.7199477 and https://doi.org/10.6084/m9.figshare.7199510.v1  
- License: CC-BY 4.0

@@

## Competing interests
No competing interests were disclosed.

@@

#### Figure 3
Taxonomy barplot as shown in the `ranacapa` Shiny app. Taxonomy is shown at the Order level in this figure; in the app, users can choose the taxonomic level to show in the barplot.
```{r}
fig3
```
